ftprime: Recording the pedigree: efficient simulation of whole genomes

### 2017-03-03 tssim.tex
algorithm, begun

These efficient methods are not available if either key assumption of coalescent theory --
neutrality or random mating -- are broken.
(Extensions can be made for only a few loci under selection \citep{hudson,krone}.)
Forwards-time simulation, however, is burdened by recording and passing on entire genomes,
which typically have tens to hundreds of millions of varying sites.

@@

>>>>

These efficient methods are not available if either key assumption of coalescent theory --
neutrality or random mating -- are broken.
(Extensions can be made for only a few loci under selection \citep{hudson,krone}.)
Forwards-time, individual-based simulation, however, is burdened by recording and passing on entire genomes,
which typically have tens to hundreds of millions of varying sites.
Such simulations must explicitly record any non-neutral genetic variants,
because these affect the demographic process.
However, neutral variants, by definition, do not, 
and so if the demographic history is recorded
they can be simulated afterwards.

Wrap-up paragraph.  
This is what we do, which lets us both record history
and be more efficient.

@@

Since msprime provides a number of tools for manipulating tree sequences,
adding mutations,
and producing output (e.g., VCF files)
we need only determine a method for properly recording history and passing it to msprime.

\subsection{msprime: Tree sequences and coalescence records}

First we recall the data storage methods used by msprime \citep[see][for more details]{msprime}.

To record the history and identity of a set of sampled genomes,
it is only necessary to record commmon ancestry (or, \emph{coalescence}) events --
most recent common ancestors of distinct segments of sampled genomes --
and the identity and location of any mutations that these most recent common ancestors carry
that are not shared by any other recorded ancestor.
\plr{(not quite right)}

The topology of the ARG
is recorded by msprime as a list of Coalescence Records, which are tuples of the form
\texttt{(left,right,parent,children)},
which indicates that the individuals listed in \texttt{children}
have inherited the segment of genome between \texttt{left} and \texttt{right}
from the ancestor \texttt{parent}.
Individuals are identified by integers,
and there is also a table recording the \emph{birth time} of each individual.

For a set of coalescence records to be consistent,
it is required that:
\begin{enumerate}
    \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
    \item The set of intervals on which individual $a$ is a child must be disjoint, for every $a$.
    \item The set of intervals on which individual $a$ is a parent must be disjoint, for every $a$.
\end{enumerate}
These ensure that the coalescence records unambiguously specify a subset of the genealogy at any location on the genome,
but do not ensure that the entire history of the samples is specified.
Furthermore, to allow for efficient algorithms, msprime also requires that
\begin{enumerate} \setcounter{enumi}{3}
    \item The samples must be numbered \texttt{0,...,n-1}, and the smallest non-sampled label must be an internal node.
        \plr{perhaps remove this one}
    \item The list of \texttt{children} in a coalescence record must be sorted.
    \item Records must be sorted in nondecreasing order by birth time of the \texttt{parent}.
    \item Birth times must be strictly greater than zero.
\end{enumerate}

\subsection{Recording history}

Since we do not know from which ancestors
the eventual genetic samples will inherit,
to record the history of the samples
we must record the results of all reproduction events -- 
the entire ARG.
To do this, and conform with the format above, we maintain
an ordered dictionary \texttt{arg},
with one entry for each parent,
whose entries are of the form
\texttt{id:(time,records)},
where \texttt{id} is the integer ID of the parent,
\texttt{time} is the birth time of the parent,
and \texttt{records} is a list of nonoverlapping coalescence records, sorted by left endpoint.
We update this structure each time a new individual is born.
Suppose that individual $I$ is born at time $t$ to parents $p_0$ and $p_1$,
with recombination breakpoints $0=x_0 < x_1 < x_2 < \cdots < x_n \le 1$,
so that $I$ inherits the genomic segment $[0,x_1)$ from $p_0$,
the segment $[x_1,x_2)$ from $p_1$, and so forth (here $n \ge 1$).
When this occurs, we:
\begin{enumerate}
    \item append a new entry \texttt{I:(t,[])} to the list \texttt{arg}; and
    \item for each $1 \le k \le n$, merge the new record \texttt{(x[k-1],x[k],p[k mod 2],(I,))}
        into the existing records in `arg[p[k mod 2]]`.
\end{enumerate}
We need to merge these records
so as to retain sortedness described above.
